---
title: "Project 1 - eBird Data"
author: "Riley McDonnell and Savannah Weaver"
date: "2/7/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Get Data
This data is of citizen science observations of the Black Capped Chickadee (*Poecile atricapillus*) and Northern Cardinal (*Cardinalis cardinalis*). This and other similar datasets can be requested online (eBird Basic Dataset. Version: EBD_relDec-2019. Cornell Lab of Ornithology, Ithaca, New York. Dec 2019.). 
*Note that the dataset is very large so we only read in the first 1000 rows otherwise it would take way too long to load in.*

```{r}
# read in whole dataset
#bcc<- read.delim("./ebd_black_capped_chickadee/bcc_data.txt", header = TRUE, sep = "\t", dec = ".", nrows=1000)
#nc<- read.delim("./ebd_northern_cardinal/nc_data.txt", header = TRUE, sep = "\t", dec = ".", nrows=1000)

# put into one dataframe
#together<- rbind(bcc,nc)

# write to new csv to make faster in future
#write.csv(together, file = "eBird_data.csv")
eBird.df<- read.csv("eBird_data.csv")
```

